---
title: "Regex practice"
author: "Cedric Arisdakessian"
date: "2022-09-07"
fontsize: 10pt
output:
  beamer_presentation:
    theme: "Antibes"
    slide_level: 2
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "pdf/") })
editor_options: 
  markdown: 
    wrap: 90
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include=TRUE)
```

# Before we begin

## Syntax review

-   `"."`: any character.
-   `x|y`: x OR y
-   `[xyz]`: x OR y OR z
-   `[x-y]`: any character in range (numbers or letters)
-   `[^xyz]`: anything but x, y or z
-   `"^"`: matches the beginning of the text.
-   `"$"`: matches the end of the text
-   `"*"`: 0+ repetitions
-   `"+"`: 1+ repetitions
-   `{n,m}`: between n and m repetitions

## Some R functions that use REGEX

- Find pattern in text: \
  `grep/grepl(pattern, text)`
- Replace pattern in text \
  `gsub(pattern, replacement, text)`

## Some bash commands that use REGEX

- Find pattern in file/stream \
  `grep [PATTERN] FILE`
- Replace pattern in file/stream \
  `sed 's/[PATTERN]/[REPLACEMENT]/g' FILE`
- And a few more

## Remarks

- In complex cases, it's better to make multiple regex queries than try to account for everything in a single expression.
- Think about all the cases you could encounter for YOUR data. Don't try to be too broad
- But make sure you're specific enough: you don't want to match things that shouldn't (see example)

# Exercises
## Word combination (question)
Find any (realistic) combination of two words in a table.

Ex: 

- "Little Pond" 
- "little_pOnd"

***

## Word combination (solution)

```{r}
ponds <- c("Little Pond", "little_pOnd ", "little.pond", 
           "little-pond", "pond")

# General expression
regex <- " *[Ll][Ii][Tt]{2}[Ll][Ee][^A-Za-z0-9][Pp][0Oo][Nn][Dd] *"
grepl(regex, ponds)

# Simpler
regex <- " *little[^A-Za-z0-9]pond *"
grepl(regex, ponds, ignore.case=T)
```
Note: if we were using `grep` on a file, we can ignore case with the `-i` option

## Missing values (question)

Handle missing values in a csv formatted file:

1. Find lines with missing entries
2. Remove any missing entry labels

How do we identify missing values in a csv file?

- Special values ("NA", "NaN", etc.)
- Empty field (",,")

## Missing values (solution)

```{r}
content <- c(
  "H,1,1,1,1",
  "He,2,NA,10,20",
  "Na,2,1,10,#VALUE",
  "Cu,,2,5,10",
  ",5,10,2,1"
)

regex <- "(N[Aa]N)|NA|(#VALUE)|(,,)|(^,)|(,$)"
grepl(regex, content)
```
Note: Careful not to match Na (since in this case, it stands for Sodium)

***
## Missing values (solution)

Let's try to do question 2) directly
```{r}
for (line in content) {
  print(sprintf(
    "%s -> %s", line, gsub(regex, "", line)
  ))
}
```
***

Solution: just don't match empty fields
```{r}
gsub("(N[Aa]N)|NA|(#VALUE)", "", content)
```

## Phone numbers (question)

How would you find US phone numbers in a text?

## Phone numbers (answer)

How would you find US phone numbers in a text?

```{r}
numbers <- c(
  "(808)-000-1111",
  "000-111-2222",
  "0001112222",
  "+33 (0)6 11 22 33 44",
  "808-11-1111"
)

country_code <- "(\\+[0-9]{1,3})?" # optional
area_code <- "\\(?[0-9]{3}\\)?"
regex <- paste(country_code, area_code, "[0-9]{3}", "[0-9]{4}", 
               sep="[- ]?")

grepl(regex, numbers)
```

## Passwords (question)

- Check if a password contains at least one capital letter
- Check if a password contains at least one special character
- Both?
